---
title: "Create new generators"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Create new generators}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
set.seed(1623L)
RcppRNG::dqset_seed(1623L)
```

```{Rcpp, ref.label=knitr::all_rcpp_labels(), include=FALSE}
```

## Summary

This document describes how you can extend `RcppRNG` with your own generator classes for random 
numbers. 

As an example, we will create an implementation of a generator for the Erlang distribution, see [https://en.wikipedia.org/wiki/Erlang_distribution](https://en.wikipedia.org/wiki/Erlang_distribution). 
Let \( k \in \mathbb{N} \) and \( \lambda > 0 \). The \( \mathrm{Erlang}{(k, \lambda)} \) distribution
has the stochastic representation 
\[
  \sum_{j=1}^{k} E_j ,
\]
where \( E_1, \ldots, E_k \) are iid exponentially distributed with rate \( \lambda \). 


## Setup 

First, you have to tell `Rcpp` that you are using `c++11` and that your code links to the 
packages `RcppRNG` (for obvious reasons) as well as `dqrng`, `BH`, and `sitmo` (to use the RNG from 
`dqrng`). 

```{Rcpp Erlang-setup, eval=FALSE}
#include <Rcpp.h>
// [[Rcpp::plugins("cpp11")]]
// [[Rcpp::depends(RcppRNG,dqrng,BH,sitmo)]]

#include <RcppRNG.hpp>
```


## Implementation

### The distribution

```{Rcpp Erlang-distribution, eval=FALSE}
namespace RcppRNG {
class ErlangDistribution : public UnivariateDistribution {
public:
  ErlangDistribution() :
      shape_(1), rate_(1.) {}
  ErlangDistribution(int shape, double rate) :
      shape_(shape), rate_(rate) {}
  
  int shape() const;
  double rate() const;
private:
  int shape_;
  double rate_;
}; // ErlangDistribution

int ErlangDistribution::shape() const {
  return shape_;
}

double ErlangDistribution::rate() const {
  return rate_;
}
} // RcppRNG
```

### The generator

```{Rcpp Erlang-generator, eval=FALSE}
namespace RcppRNG {
template<typename T>
class ErlangGenerator : public Generator<T, double, ErlangDistribution> { 
public:
  ErlangGenerator();
  ErlangGenerator(const ErlangDistribution& param);
  
  inline double operator()() const;
private:
  ErlangDistribution param_;
  ExpGenerator<T> exp_generator_;
}; // ErlangGenerator

template<typename T>
ErlangGenerator<T>::ErlangGenerator() : 
    param_(), exp_generator_(ExpGenerator<T>(ExpDistribution<double>(1. / param_.rate()))) {
}

template<typename T>
ErlangGenerator<T>::ErlangGenerator(const ErlangDistribution& param) : 
    param_(param), exp_generator_(ExpGenerator<T>(ExpDistribution<double>(1. / param_.rate()))) {}

template<typename T>
inline double ErlangGenerator<T>::operator()() const {
  double out = 0.;
  for (int j=1; j<=param_.shape(); j++)
    out += exp_generator_();

  return out;
}
} // RcppRNG
```

### `Rcpp`-wrapper

```{Rcpp drop-in-functions, eval=FALSE}
using namespace Rcpp;

//' @export
// [[Rcpp::export(rng=false)]]
NumericVector Rcpp_rerlang_RcppRNG(R_xlen_t n, int shape = 1, double rate = 1.) {
  NumericVector out(no_init(n));
  RcppRNG::ErlangDistribution param(shape, rate);
  RcppRNG::ErlangGenerator<RcppRNG::RcppRNG> gen(param);
  std::generate(out.begin(), out.end(), gen);

  return out;
}

//' @export
// [[Rcpp::export(rng=false)]]
NumericVector Rcpp_rerlang_DQRNG(R_xlen_t n, int shape = 1, double rate = 1.) {
  NumericVector out(no_init(n));
  RcppRNG::ErlangDistribution param(shape, rate);
  RcppRNG::ErlangGenerator<RcppRNG::DQRNG> gen(param);
  std::generate(out.begin(), out.end(), gen);

  return out;
}
```

## Test 

```{r test-Erlang-generator}
Rcpp_rerlang_RcppRNG(10, 3, 0.5)
Rcpp_rerlang_DQRNG(10, 3, 0.5)
```

